---
title: "Pacific giant salamanders (Dicamptodon tenebrosus) in Mack Creek (near Blue River, Oregon): Size and channel location in old growth and previously clear cut forest sections"
subtitle: "ESM 206 Assignment 5"
author: "Kaitlyn Lanfield and Ashley Stroud"
date: "Due: Decemeber 4, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor) # for clean names function
library(directlabels) # for direct labels on line graph
library(kableExtra) # to create table
library(effsize)
library(ggbeeswarm)
library(car)

#original dataframe
mack_creek_vertebrates_og <- read_csv("mack_creek_vertebrates.csv") %>% 
  clean_names()

#pacific salamander only DF
pacific_salamander <- mack_creek_vertebrates_og %>% 
  filter(species == "DITE") %>% 
  mutate(forest_type = case_when(section == "CC" ~ "Clear cut",
                                 section == "OG" ~ "Old growth" )) %>%
  mutate(channel_type = case_when(unittype == "C" ~ "Cascade",
                                  unittype == "P" ~ "Pool",
                                  unittype == "SC" ~ "Side channel"))
  
  

```


<br>

###Introduction



<br>

###Data and Methods

####Results A: Visually compare annual salamander counts in old growth (OG) and clear cut (CC) sections of Mack Creek.
```{r, fig.cap= "***Figure 3.*** *Annual pacific salamander counts in old growth and clear cut sections of Mack Creek (1993 - 2017).*"}

#pacific salamander counts df
pacific_salamander_count <- pacific_salamander %>% 
  count(year, forest_type, name = "count")

#plot 
years_growth_plot <- ggplot(pacific_salamander_count, aes(x = year, y = count)) +
  geom_line(aes(color = forest_type)) +
  geom_point() +
  scale_x_continuous(limits = c(1993, 2017), expand = c(0,0), breaks = seq(0, 2017, 5)) +
  scale_y_continuous(limits = c(0, 400), expand = c(0,0), breaks = seq(0, 380, 50)) +
  scale_color_manual(values = c("green", "forestgreen"), labels = c("Clear cut", "Old growth")) +
  labs(x = "Year",
       y = "Abundance",
       color = "Forest type")

years_growth_plot


#we can see that they are generally the same trend, many of the years clear cut is less, but overall trending in numbers other than that steep decline

```


####Results B: Table of 2017 salamander counts by channel classification (pool, cascades and side-channel) in old growth (OG) and clear cut (CC) sections of Mack Creek.
```{r}

#ps is for pacific salamander
ps_2017_channel_count <- pacific_salamander %>% 
  filter(year == "2017") %>% 
  count(channel_type, forest_type, name = "count") %>% 
  na.omit()

# create table
ps_2017_channel_table <- ps_2017_channel_count %>% 
  pivot_wider(names_from = channel_type, values_from = count) 


ps_2017_channel_props <- ps_2017_channel_table %>% #wide format table 
  adorn_percentages(denominator = "row") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns(position = "front") %>% #if i convert to proportins lose sample sixze, so add back in with n's, and put in front 
  kable() %>% 
  kable_styling()

ps_2017_channel_props



```

####Results C: Using the 2017 counts found in Results B above, answer: is there a significant difference in where in the channel Pacific giant salamanders are located (pool, cascade or side channel) between the two sections (old growth and clear cut)?
```{r}
# Make this a contingency table that **only contains the actual counts** to perform chi sqaure

chi_ps_2017_channel_count <- ps_2017_channel_table %>% 
  select(-forest_type)

chi_ps_2017_channel_count

chi_test_ps_2017_channel <- chisq.test(chi_ps_2017_channel_count)
chi_test_ps_2017_channel

```


####Results D. Compare weights of Pacific giant salamanders in clear cut and old growth forest sections of the creek in 2017.
```{r}
colnames(pacific_salamander)
ps_2017_weight <- pacific_salamander %>% 
  filter(year == "2017") %>% 
  select(year, forest_type, channel_type, weight)

#exploratory data viz

ggplot(ps_2017_weight, aes(x = weight)) +
  geom_histogram(aes(fill = forest_type))

#even though we see outliers and the distribution are not normal, bc we have such high sample sizes we can use the central limit theorem


```


```{r}
#t-tests

cc_weight <- ps_2017_weight %>% 
  filter(forest_type == "Clear cut")

og_weight <- ps_2017_weight %>% 
  filter(forest_type == "Old growth")

ps_weight_ttest <- t.test(cc_weight$weight, og_weight$weight)
ps_weight_ttest

# there is not a sig diff between mean salamnder weights in 2017 between clear cut and old growth 

```


####Results E. Compare weights of Pacific giant salamanders in pools, cascades and side-channels of Mack Creek in 2017.
```{r}
ps_2017_channel_weight <- ps_2017_weight %>% 
  select(year, channel_type, weight) %>% 
  na.omit()

#visual exploration
ggplot(ps_2017_channel_weight, aes(x = weight)) +
  geom_histogram(aes(fill = channel_type)) +
  facet_wrap(~channel_type)
# here we see positively skewed data, outliers, while means could be used due to CLT, it would not be the most appropriate representation of central tendency 

```




```{r}

ps_2017_channel_summary <- ps_2017_channel_weight %>% 
  group_by(channel_type) %>% 
  summarise(
    sample_size = n(),
    mean_weight = mean(weight),
    sd_weight = sd(weight),
    se_weight = sd(weight)/ sqrt(n())
    )
ps_2017_channel_summary

```


```{r}



ps_2017_channel_plot <- ggplot() +
  geom_beeswarm(data = ps_2017_channel_weight, 
                aes(x = channel_type, y = weight), 
                size = 1,
                alpha = 0.6,
                color = "gray60") +
  geom_point(data = ps_2017_channel_summary, 
             aes(x = channel_type, y = mean_weight)) +
  geom_errorbar(data = ps_2017_channel_summary,
                aes(x = channel_type, 
                    ymin = mean_weight - se_weight,
                    ymax = mean_weight + se_weight),
                width = 0.1) +
  coord_flip()

ps_2017_channel_plot




```


```{r}
#Levene's test (hypothesis test for equal variance):

leveneTest(weight ~ channel_type, data = ps_2017_channel_weight)

#H0: variances are equals, there is no difference in variance 
#HA: variances are not equal, there is a difference in variance

# given that p greater than 0.05, we retain the null that variances are equal or not sig different, so we can proceed with an anova which has the assumption that variances are equal 

ps_aov <- aov(weight ~ channel_type, data = ps_2017_channel_weight)
summary(ps_aov)
# here we see that it is sig, so move forward with tukeys, also need to comment the p value here

TukeyHSD(ps_aov)
# only sig is side channel to the pool

#This creates a summary of all the outputs from our ANOVA so we can reference it 

ps_aov_outputs <- unlist(summary(ps_aov))
ps_aov_outputs

```

```{r}
# Now we want to do effect sizes on each of the populations regardless of their significant differences

#Here we will split them each into different dataframes

cascade_weight <- ps_2017_channel_weight %>% 
  filter(channel_type == "Cascade")

pool_weight <- ps_2017_channel_weight %>%
  filter(channel_type == "Pool")

side_channel_weight <- ps_2017_channel_weight %>%
  filter(channel_type == "Side-channel")

#Here we will test the effect sizes of each channel type interactions

#Effect size of cascade salamander weights and pool salamander weights

cascade_pool_ef <- effsize::cohen.d(cascade_weight$weight, pool_weight$weight)
cascade_pool_ef

#Effect size is negligible which is not surprising because the difference was insignificant in our Tukey's HSD

#Effect size of side_channel salamander weights and cascade salamander weights

cascade_side_channel_ef <- effsize::cohen.d(cascade_weight$weight, side_channel_weight$weight)
cascade_side_channel_ef

#Effect size is small

#Effect size of side_channel salamander weights and pool salamander weights

pool_side_channel_ef <- effsize::cohen.d(pool_weight$weight, side_channel_weight$weight)
pool_side_channel_ef

#Effect size is small

```





####Questions
-how does this beeswarm work for a visualization, especially since we inverted it
-what would be an example of using confidence interval
-im confused how we get standard error when we dont know the pop sd? or do we? 
-effect size w annova, not f distribution 

