---
title: "Pacific giant salamanders (Dicamptodon tenebrosus) in Mack Creek (near Blue River, Oregon): Size and channel location in old growth and previously clear cut forest sections"
subtitle: "ESM 206 Assignment 5"
author: "Kaitlyn Lanfield and Ashley Stroud"
date: "Due: Decemeber 4, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor) # for clean names function
library(directlabels) # for direct labels on line graph
library(kableExtra) # to create table
library(effsize)
library(ggbeeswarm)
library(car)

#original dataframe
mack_creek_vertebrates_og <- read_csv("mack_creek_vertebrates.csv") %>% 
  clean_names()

#pacific salamander only DF
pacific_salamander <- mack_creek_vertebrates_og %>% 
  filter(species == "DITE") %>% 
  mutate(forest_type = case_when(section == "CC" ~ "Clear cut",
                                 section == "OG" ~ "Old growth" )) %>%
  mutate(channel_type = case_when(unittype == "C" ~ "Cascade",
                                  unittype == "P" ~ "Pool",
                                  unittype == "SC" ~ "Side channel"))
  
  

```


<br>

### Introduction



<br>

### Data and Methods

#### A: Changes in annual salamander abundance in old growth and clear cut sections of Mack Creek.
Abundance in pacific salamanders has increased overall in both clear cut and old growth forests from 1993 to 2017. While the two salamander populations share a similar trend of incline and decline in counts throughout the years, the counts of salamanders in clear cut forests are generally lower than those inhabiting old growth forests. Interestingly, this trend switched in 2014 (Figure 3.). 

```{r, fig.cap= "***Figure 3.*** *Annual pacific salamander counts in old growth and clear cut sections of Mack Creek (1993 - 2017).*"}

#pacific salamander counts df
pacific_salamander_count <- pacific_salamander %>% 
  count(year, forest_type, name = "count")

#plot 
years_growth_plot <- ggplot(pacific_salamander_count, aes(x = year, y = count)) +
  geom_line(aes(color = forest_type)) +
  geom_point(aes(color = forest_type)) +
  scale_x_continuous(limits = c(1993, 2017), expand = c(0,0), breaks = seq(0, 2017, 5)) +
  scale_y_continuous(limits = c(0, 400), expand = c(0,0), breaks = seq(0, 380, 50)) +
  scale_color_manual(values = c("green", "forestgreen"), labels = c("Clear cut", "Old growth")) +
  labs(x = "Year",
       y = "Abundance",
       color = "Forest type")+
  theme_minimal()

years_growth_plot


#we can see that they are generally the same trend, many of the years clear cut is less, but overall trending in numbers other than that steep decline

```


#### Results B: Table of 2017 salamander counts by channel classification (pool, cascades and side-channel) in old growth (OG) and clear cut (CC) sections of Mack Creek.
```{r}

#ps is for pacific salamander
ps_2017_channel_count <- pacific_salamander %>% 
  filter(year == "2017") %>% 
  count(channel_type, forest_type, name = "count") %>% 
  na.omit()

# create table
ps_2017_channel_table <- ps_2017_channel_count %>% 
  pivot_wider(names_from = channel_type, values_from = count) 


ps_2017_channel_props <- ps_2017_channel_table %>% 
  adorn_percentages(denominator = "row") %>% 
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns(position = "front") %>% 
  kable(col.names = c("Forest type", "Cascade", "Pool", "Side-channel")) %>% 
  kable_styling() %>% 
  add_header_above(c("Channel type " = 4))

ps_2017_channel_props


```

#### Results C: Using the 2017 counts found in Results B above, answer: is there a significant difference in where in the channel Pacific giant salamanders are located (pool, cascade or side channel) between the two sections (old growth and clear cut)?
```{r}
# Make this a contingency table that **only contains the actual counts** to perform chi sqaure

chi_ps_2017_channel_count <- ps_2017_channel_table %>% 
  select(-forest_type)

chi_ps_2017_channel_count

chi_test_ps_2017_channel <- chisq.test(chi_ps_2017_channel_count)
chi_test_ps_2017_channel

```

There is no significant effect of forest type (clear cut/old growth) on where in the channel (cascade/pool/side-channel) pacific salamanders are found. ($\chi$^2^(`r chi_test_ps_2017_channel$parameter`) = `r round(chi_test_ps_2017_channel$statistic,2)`, *p* = `r round(chi_test_ps_2017_channel$p.value, 2)`). The percentage of salamanders found in cascades and side-channels differed by less than 5% between clear cut and old growth forests. Percentages of salamanders found in pools differed slightly more, but still by less than 10%. 

#### Results D. Mean weights of Pacific giant salamanders in clear cut and old growth forest sections of the creek in 2017. or effects of forest type on salamander weight??
```{r}
colnames(pacific_salamander)
ps_2017_weight <- pacific_salamander %>% 
  filter(year == "2017") %>% 
  select(year, forest_type, channel_type, weight)

#exploratory data viz

ggplot(ps_2017_weight, aes(x = weight)) +
  geom_histogram(aes(fill = forest_type))

#even though we see outliers and the distribution are not normal, bc we have such high sample sizes we can use the central limit theorem


```

```{r}
#we would like compare mean salamander weight in 2017 in clear cut and old growth forests. we first create data fromes for weights in each forest in 2017 only. 

#clear cut 2017 weights dataframe
cc_weight <- ps_2017_weight %>% 
  filter(forest_type == "Clear cut")

#old growth 2017 weights data frame
og_weight <- ps_2017_weight %>% 
  filter(forest_type == "Old growth")

#running a t-test and effect size to statisically evaluate the difference in weights
ps_weight_ttest <- t.test(cc_weight$weight, og_weight$weight)
ps_weight_ttest
# t = 1.6669, df = 692.79, p-value = 0.09599, mean of x= 7.77, mean of y = 6.58, not a significant difference using a CI of 5%

#effect size
ps_weight_eff <- cohen.d(cc_weight$weight, og_weight$weight, na.rm = TRUE)
ps_weight_eff
#effect size is negligible

# there is not a sig diff between mean salamnder weights in 2017 between clear cut and old growth 

```
Weights were compared of pacific giant salamanders in old growth forests to those in clear cut forests in 2017 only. There was not a significant difference in weights of salamanders between the two forest types (Welch Two Sample t-test, t = `r round(ps_weight_ttest$statistic,2)`, df = `r round(ps_weight_ttest$parameter,2)`, p = `r round (ps_weight_ttest$p.value,3)`).  




#### Results E. Compare weights of Pacific giant salamanders in pools, cascades and side-channels of Mack Creek in 2017.
```{r}
ps_2017_channel_weight <- ps_2017_weight %>% 
  select(year, channel_type, weight) %>% 
  na.omit()

#visual exploration
ggplot(ps_2017_channel_weight, aes(x = weight)) +
  geom_histogram(aes(fill = channel_type)) +
  facet_wrap(~channel_type)
# here we see positively skewed data, outliers, while means could be used due to CLT, it would not be the most appropriate representation of central tendency 

```




```{r}
# create a summary DF for each of the channel types to be used for plotting  
ps_2017_channel_summary <- ps_2017_channel_weight %>% 
  group_by(channel_type) %>% 
  summarise(
    sample_size = n(),
    mean_weight = mean(weight),
    sd_weight = sd(weight),
    se_weight = sd(weight)/ sqrt(n()),
    var_weight = var(weight)
    )
ps_2017_channel_summary

#largest variance is 185.58, smallest variance is 68.31, largest is less than 4x, variances are likely close enough

```


```{r}



ps_2017_channel_plot <- ggplot() +
  geom_beeswarm(data = ps_2017_channel_weight, aes(x = channel_type, y = weight, color = channel_type), size = 1, alpha = 0.6) +
  scale_color_manual(values = c( "#3e64ff","#5edfff","#b2fcff")) +
  scale_y_continuous(limits = c(0, 105), expand = c(0,0), breaks = seq(0, 100, 25))+
  theme_minimal()+
  geom_point(data = ps_2017_channel_summary, aes(x = channel_type, y = mean_weight), size = 0.8)+
  geom_errorbar(data = ps_2017_channel_summary, aes(x = channel_type,
                                                    ymin = mean_weight - se_weight,
                                                    ymax = mean_weight + se_weight),
                width = 0.1)+
  theme(legend.position = "none") +
  theme(axis.title.y = element_blank())+
  labs(y = "Pacific salamander weight(g)")+
  coord_flip()

ps_2017_channel_plot


#2E86C1
#48C9B0
#DC7633


```


```{r}
#Statistical analysis of weights between the three channel types:


#----------------------------Levene's test------------------------------------# 
#(hypothesis test for equal variance) for the weights of three channel types:

leveneTest(weight ~ channel_type, data = ps_2017_channel_weight)

#H0: variances are equals, there is no difference in variance 
#HA: variances are not equal, there is a difference in variance

# given that p greater than 0.05, we retain the null that variances are equal or not sig different, so we can proceed with an anova which has the assumption that variances are equal. We expected this since we saw that the largest was less than 4x in the summary table df we created. 
#-------------------------------------------------------------------------------# 


#----------------------------ANOVA----------------------------------------------#
ps_aov <- aov(weight ~ channel_type, data = ps_2017_channel_weight)
summary(ps_aov)
# here we see that it is sig, so move forward with tukeys, also need to comment the p value here
#-------------------------------------------------------------------------------# 


#--------------------------Post-hoc analysis------------------------------------# 
TukeyHSD(ps_aov)
# only sig is side channel to the pool

#-------------------------------------------------------------------------------# 


#-------------------------ANOVA summary----------------------------------------# 
ps_aov_outputs <- unlist(summary(ps_aov))
ps_aov_outputs #This creates a summary of all the outputs from our ANOVA so we can reference it #-------------------------------------------------------------------------------# 

```

```{r}
# Now we want to do effect sizes on each of the populations regardless of their significant differences

#----------------------------Channel DF's------------------------------------------
#Here we will split them each into different dataframes so that we are able to use the dateframe for effsize test

cascade_weight <- ps_2017_channel_weight %>%  
  filter(channel_type == "Cascade")

pool_weight <- ps_2017_channel_weight %>%
  filter(channel_type == "Pool")

side_channel_weight <- ps_2017_channel_weight %>%
  filter(channel_type == "Side-channel")
#--------------------------------------------------------------------------------


#-------------------------------Effect size---------------------------------------
#Here we will test the effect sizes of each channel type interactions

#cascade salamander weights - pool salamander weights
cascade_pool_ef <- effsize::cohen.d(cascade_weight$weight, pool_weight$weight)
cascade_pool_ef
#Effect size is negligible which is not surprising because the difference was insignificant in our Tukey's HSD


#Eside_channel salamander weights -cascade salamander weights
cascade_side_channel_ef <- effsize::cohen.d(cascade_weight$weight, side_channel_weight$weight)
cascade_side_channel_ef
#Effect size is small


#side_channel salamander weights - pool salamander weights
pool_side_channel_ef <- effsize::cohen.d(pool_weight$weight, side_channel_weight$weight)
pool_side_channel_ef
#Effect size is small

#-------------------------------------------------------------------------------# 
```





References
Gregory S. V. 2016. Aquatic Vertebrate Population Study in Mack Creek, Andrews Experimental Forest, 1987 to present. Environmental Data Initiative. https://doi.org/10.6073/pasta/5de64af9c11579266ef20da2ff32f702. Dataset accessed 11/19/2019.


Questions
-how do we have beeswarm not overlap
-do we need to write p(0.05) = ??
-side channel color seems a little light for my eyes
